---
title: "Chapter 4 -- Classification"
subtitle: "Lab: Linear discriminant analysis"
author: "Diogo M. Camacho"
date: "`r paste('Last updated:', format(Sys.Date(), '%Y-%m-%d'))`"
output: html_notebook
---
  

```{r libraries, echo = FALSE}
# libraries and data
library(ISLR)
library(dplyr)
library(ggplot2)
library(GGally)
library(MASS)
library(caret)
library(readr)  
library(corrplot)
```

This is an exploration of the `Stock Market` data to perform classification on these data.

## 4.6.1 The `Stock Market` data
The stock market data consists of percentage return for the S&P 500 stock index over 1,250 days. For any given date, the percentage return for each of the previous 5 trading days was calculated, and named `Lag1` throu `Lag5` in the data set. Additionally, `Volume` (number of shares traded on the previous day, in billions), `Today` (the percentage return on the date in question), and `Direction` (wether the market was up or down on this date)


```{r}
smarket <- ISLR::Smarket
```

```{r}
head(smarket)
```

```{r}
dim(smarket)
```

```{r}
summary(smarket)
```

```{r}
# pairs(smarket)
smarket %>%
  as_tibble %>%
  GGally::ggpairs(columns = c("Lag1", "Lag2", "Lag3", "Lag4", "Lag5", "Volume", "Today"), alpha = 0.5, progress = FALSE)
```

Now I will use `corrplot` to display the correlation between all variable in the data set.

```{r}
smarket %>% 
  dplyr::select(., -Direction) %>%
  cor() %>%
  corrplot::corrplot.mixed(., lower.col = "black")
```

We see that the only relevant correlation is between `Year` and `Volume`, which is equal to `r cor(smarket[, -9])["Year", "Volume"]`. We can plot the data to see this relationship:

```{r}
smarket %>% 
  dplyr::select(., -Direction) %>% 
  tibble::add_column(.,  seq_days = seq(1, nrow(smarket))) %>% 
  ggplot(aes(x = seq_days, y = Volume)) + 
  geom_point(color = "red", alpha = 0.5, size = 3) + 
  labs(y = "Volume", x = "Days") + 
  theme_bw()
```

## 4.6.3 Linear Discriminant Analysis
To fit an LDA model we use the `lda()` function from the `MASS` package. We will also generate a training set that reports only the stock market data prior to 2005.

```{r training_data}
train <- (smarket$Year < 2005)
```

Now we fit the linear discriminant analysis model:

```{r lda_model}
lda_model <- lda(Direction ~ Lag1 + Lag2, data = smarket, subset = train)

lda_model
```

We can now make predictions on the unseen data (data from 2005) using the LDA model we just trained:

```{r lda_preds}
lda_prediction <- predict(object = lda_model, newdata = smarket[!train, ]) # <-- train is only IDs so I can do this
```

```{r pred_res}
table(lda_prediction$class, smarket$Direction[!train])
```

